What does it take to create with domain-appropriate tools?

### 2020-05-14 convivial-2020/manuscript.tex
No more Pandoc tilde BS, further rework

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it\footnote{Besides
  that, the One-Size-Fits-All{} approach is perhaps unavoidable for
  static print media.}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-html.png}
  \caption{Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of Id{} objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the Id{}/HTML implementation \cite{orom-html},
in which they take the form of HTML tables.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In Id{}/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer \cite{boxer}. This was a
significant departure from the table representation, and even though SVG
supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

@@

Most software is designed for the subset of people who have a colour
display they can perceive. So right away it is going to require ways to
draw coloured shapes. There are usually libraries for this (though note:
not part of the language), but some only provide \emph{immediate mode}:
commands to instantaneously rasterise pixels to a buffer. This is not
enough for modern software, as we often expect animation, or at least to
see things change as we interact. Most often we wish to see \emph{small
changes} to the \emph{same} shapes, rather than completely different
shapes altogether; the reification that this requires to persist between
frames, is known as \emph{retained mode.}

@@

The SVG tree has the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this would boil down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see Section \ref{context-appropriate-ontologies}.

@@

It starts with a consideration of translational rigidity. In its most
primitive form, this is a relation between two points. Thus it is
natural to draw a line or ``rod'' between them. It seems that this rod
transmits changes in one of its endpoints directly to the other
endpoint. Since they both feel the same deltas, the displacement vector
between them is preserved (Figure \ref{fig:delta-transmission}).

@@

Resizing of boxes could be achieved through rods that stay horizontal or
vertical. In the language of ``small differences'' spoken by the
live-state infrastructure, this is expressed as a rod ``transmitting
deltas'' in the vertical and horizontal, ``absorbing'' the other
component into itself (Figure \ref{fig:rods}). Mirroring a DOM rect to
these rods is as simple as subscribing its width and height to
horizontal and vertical rods' \texttt{length} Observables. This way,
boxes can be resized from whatever corner is convenient.

@@

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
vector subtraction, which can then be propagated as local co-ordinates
to children. It is nice to express the relation (as well as see it!)
this way (Figure \ref{fig:rods}).

@@

Each API has its own conventions, including a way of naming and
structuring expressions---an ontology \cite{crit-semprola}. The
One-Size-Fits-All{} approach is exemplified in such interfaces. For
example, in SVG we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height---and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

For example, I find it natural to resize boxes by dragging any of their
four corners, so I wanted this in Id{}/SVG. In this context, a
``rectangle'' is \emph{seen as} four points: top-left, top-right,
bottom-left, bottom-right. Obviously this is not a \emph{minimal}
description, since given e.g.~the top-left and bottom-right, the other
two points can be inferred. But one way or another, to be able to drag
any of them, all four points must be present at \emph{some} level. This
alternative ontology was polyfilled in the form of a ``rect controls''
class that can be attached manually to any SVG
\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
are subscribed to the top-left Point\footnote{In the usual case where
  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
  top-left instead controls the parent
  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
to Rod lengths.

@@

What would it look like to support multiple ontologies? Na"ively, there
are two answers: anticipate the possibilities ahead of time, or support
users adding their own. It is unclear if we can do better on the latter
than simply ``support polyfilling''. As for the former, anticipating the
diversity of ways someone might look at the world is generally doomed to
fail. But in the special case of fairly \emph{formalised} concepts such
as geometrical shapes or mathematics, I could suggest that simple
\emph{under-specification} is the root of the problem in SVG. Instead of
the specification explaining in English that ``\texttt{x} and \texttt{y}
are the co-ordinates of the top left-hand corner\ldots{}'', it might be
better to make these relations machine-readable or \emph{embodied} in
the API. For the rect this might look like:

@@

By contrast, after switching to the fully nestable box substrate in
Id{}/SVG, I could directly express the model I was thinking of (Figure
\ref{fig:method-boxes}). If I had stayed in HTML, there is a good chance
I would have had to polyfill in extensional functions---whether through
nested \texttt{\textless{}table\textgreater{}} elements, additional
``-'' characters or something else.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{../method-boxes.png}
  \caption{Id{}/SVG: the \texttt{vtable} box has a \texttt{methods} box, and that's where the method boxes go. \label{fig:method-boxes}}
\end{figure}

@@

Normally, this can be as simple as autosave to the filesystem. But the
Web platform is very wary of this\footnote{As anyone who learns WebGL
  can attest to, when they discover they must run a local Web server to
  provide image files for textures since any filesystem requests will be
  rejected for security.}, so the solution I turned to was manually
copying the markup in the browser's inspector (Figure
\ref{fig:html-inspector}).

@@

This argument as stated suffers by glossing over an important fact.
Maxwell's equations certainly do fit onto a T-shirt, but most people
will not be able to explain what they mean. What typically amounts to
years of study is compressed into those mathematical symbols, and the
learning material involved most certainly does \emph{not} fit on a
T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
these equations under a single symbol, \(\mathcal{M}\). \(\mathcal{M}\)
is defined as ``Maxwell's Equations are true''. Ta-da---this fits on a
coin, but good luck doing anything with it.\footnote{In fact, this is
  almost achieved by the formalism of Geometric Calculus, which reduces
  them to only one equation.}

@@

Said related ``COLAs'' paper expands on Id{}, giving it a ``structural''
role complemented by a LISP-like programming language. Section 6.1
sketches out the intended ``bootstrapping'' process, ending with a
self-sufficent, self-hosting version of the system.

@@

In Section \ref{retained-mode-vector-graphics}, I touched on how
everyone with a browser has access to a powerful vector graphics editor
(SVG) locked behind a completely inappropriate UI (the JS Console). It
is similar for 3D graphics (WebGL), and sound and music (Web Audio).
Similar observations are made about ``native'' OS apps in
\cite{prog21-dyn}. Unfortunately for that domain, the interface that
unlocks your operating system's range of functionalities---batch-mode
compilation---is quite far from the affordances of the (interactive) JS
Console.

I hope to use Id{}/SVG to ``tame'' SVG and other JavaScript APIs, with
some minimal on-demand visualisation in the large portion of the screen
next to the JS console. This is an attempt to generalise the ``rect
controls'', which allow the obvious geometric properties of a
\texttt{\textless{}rect\textgreater{}} to be directly manipulated
(Section \ref{context-appropriate-ontologies}). Giving Id{} access to
Web technologies in this way is essential for evolving it into a
\emph{fully-featured} self-changeable software environment, which could
allow as much domain-specifc representation as its user wishes.

>>>>

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams.

For example, the first acts of the running system boil down to
initialising the three or so objects. This consists of allocating
memory, interpreting it as a C struct and then filling in fields in a
mundane manner. I had great difficulty following the specifics in my
head, but when I drew tables in the style of their diagrams I readily
saw what what going on. I personally would have preferred these to have
been in the paper in the first place, but I am not necessarily
representative of those who read it\footnote{Besides that, the
  One-Size-Fits-All{} approach is perhaps unavoidable for static print
  media.}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-html-cropped.png}
  \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of Id{} objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}.
Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML
implementation \cite{orom-html}, in which they take the form of HTML
tables.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg-cropped.png}
  \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In Id{}/SVG, shown in Figure\textasciitilde{}\ref{fig:orom-svg},
obj-dicts are encoded as nested SVG
\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
of diSessa's Boxer \cite{boxer}. This was a significant departure from
the table representation, and even though SVG supports (some) nested
HTML via \texttt{\textless{}foreignObject\textgreater{}}, I actually
preferred the possibility of multiple levels of nesting.

@@

Most software is designed for the subset of people who have a colour
display they can perceive. So right away it is going to require ways to
draw coloured shapes. There are usually libraries for this (though they
aren't always \emph{standard} libraries), but some only provide
\emph{immediate mode}: commands to instantaneously rasterise pixels to a
buffer. This is not enough for modern software, as we often expect
animation, or at least to see things change as we interact. Most often
we wish to see \emph{small changes} to the \emph{same} shapes, rather
than completely different shapes altogether; the reification that this
requires to persist between frames, is known as \emph{retained mode.}

@@

The SVG tree has the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this would boil down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see
Section\textasciitilde{}\ref{context-appropriate-ontologies}.

@@

It starts with a consideration of translational rigidity. In its most
primitive form, this is a relation between two points. Thus it is
natural to draw a line or ``rod'' between them. It seems that this rod
transmits changes in one of its endpoints directly to the other
endpoint. Since they both feel the same deltas, the displacement vector
between them is preserved
(Figure\textasciitilde{}\ref{fig:delta-transmission}).

@@

Resizing of boxes could be achieved through rods that stay horizontal or
vertical. In the language of ``small differences'' spoken by the
live-state infrastructure, this is expressed as a rod ``transmitting
deltas'' in the vertical and horizontal, ``absorbing'' the other
component into itself (Figure\textasciitilde{}\ref{fig:rods}). Mirroring
a DOM rect to these rods is as simple as subscribing its width and
height to horizontal and vertical rods' \texttt{length} Observables.
This way, boxes can be resized from whatever corner is convenient.

@@

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
vector subtraction, which can then be propagated as local co-ordinates
to children. It is nice to express the relation (as well as see it!)
this way (Figure\textasciitilde{}\ref{fig:rods}).

@@

Each API has its own conventions, including a way of naming and
structuring expressions: an \emph{ontology} \cite{crit-semprola}. The
One-Size-Fits-All{} approach is exemplified in such interfaces.

In SVG, we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height---and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

I find it natural to resize boxes by dragging any of their four corners,
so I wanted this in Id{}/SVG. In this context, a ``rectangle'' is
\emph{seen as} four points: top-left, top-right, bottom-left,
bottom-right. Obviously this is not a \emph{minimal} description, since
given e.g.~the top-left and bottom-right, the other two points can be
inferred. But one way or another, to be able to drag any of them, all
four points must be present at \emph{some} level. This alternative
ontology was polyfilled in the form of a ``rect controls'' class that
can be attached manually to any SVG
\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
are subscribed to the top-left Point\footnote{In the usual case where
  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
  top-left instead controls the parent
  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
to Rod lengths.

@@

What would it look like to support multiple ontologies? Naïvely, there
are two answers: anticipate the possibilities ahead of time, or support
users adding their own. It is unclear if we can do better on the latter
than simply ``support polyfilling''. As for the former, anticipating the
diversity of ways someone might look at the world is generally doomed to
fail. But in the special case of fairly \emph{formalised} concepts such
as geometrical shapes or mathematics, I could suggest that simple
\emph{under-specification} is the root of the problem in SVG. Instead of
the specification explaining in English that ``\texttt{x} and \texttt{y}
are the co-ordinates of the top left-hand corner\ldots{}'', it might be
better to make these relations machine-readable or \emph{embodied} in
the API. For the rect this might look like:

@@

By contrast, after switching to the fully nestable box substrate in
Id{}/SVG, I could directly express the model I was thinking in. Quite
simply, the \texttt{vtable} box has a \texttt{methods} box, and that's
where the method boxes go (Figure\textasciitilde{}\ref{fig:orom-svg}).
If I had stayed in HTML, there is a good risk I would have had to
polyfill in extensional functions---whether through nested
\texttt{\textless{}table\textgreater{}} elements, additional ``-''
characters, or something else.

@@

@@

Normally, this can be as simple as autosave to the filesystem. But the
Web platform is very wary of this\footnote{As anyone who learns WebGL
  can attest to, when they discover they must run a local Web server to
  provide image files for textures since any filesystem requests will be
  rejected for security.}, so the solution I turned to was manually
copying the markup in the browser's inspector
(Figure\textasciitilde{}\ref{fig:html-inspector}).

@@

This argument as stated suffers by glossing over an important fact.
Maxwell's equations certainly do fit onto a T-shirt, but most people
will not be able to explain what they mean. What typically amounts to
years of study is compressed into those mathematical symbols, and the
learning material involved most certainly does \emph{not} fit on a
T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
these equations under a single symbol\footnote{In fact, this is almost
  achieved by the formalism of Geometric Calculus, which reduces them to
  only one equation.}, \(\mathcal{M}\). \(\mathcal{M}\) is defined as
``Maxwell's Equations are true''. Ta-da---this fits on a coin, but good
luck doing anything with it.

@@

Said related ``COLAs'' paper expands on Id{}, giving it a ``structural''
role complemented by a LISP-like programming language.
Section\textasciitilde{}6.1 sketches out the intended ``bootstrapping''
process, ending with a self-sufficent, self-hosting version of the
system.

@@

In Section\textasciitilde{}\ref{retained-mode-vector-graphics}, I
touched on how everyone with a browser has access to a powerful vector
graphics editor (SVG) locked behind a completely inappropriate UI (the
JS Console). It is similar for 3D graphics (WebGL), and sound and music
(Web Audio). Similar observations are made about ``native'' OS apps in
\cite{prog21-dyn}. Unfortunately for that domain, the interface that
unlocks your operating system's range of functionalities---batch-mode
compilation---is quite far from the affordances of the (interactive) JS
Console.

I intend to use Id{}/SVG to ``tame'' SVG and other JavaScript APIs, with
some minimal on-demand visualisation in the large portion of the screen
next to the JS console. This is an attempt to generalise the ``rect
controls'', which allow the obvious geometric properties of a
\texttt{\textless{}rect\textgreater{}} to be directly manipulated
(Section\textasciitilde{}\ref{context-appropriate-ontologies}). Giving
Id{} access to Web technologies in this way is essential for evolving it
into a \emph{fully-featured} self-changeable software environment, which
could allow as much domain-specifc representation as its user wishes.
